Markov Link Method for calibrating without joint measurement, including the case of destructive measurements

### 2018-05-21 paper/paper.tex
Used full nxx types for the analysis

@@

\title{Markov Link Method for calibrating without joint measurement, including the case of destructive measurements}
\author{Jackson Loper}

@@

\theoremstyle{definition}
\newtheorem{example}{Example}

@@

\begin{abstract}
A proliferation of new experimental tools has left a serious gap: calibration. Two thermometers can be calibrated against each other by simply measuring the same bodies of water with both thermometers, but the problem is much harder for many modern tools.  One common problem is that we do not have measurements from the same ``body of water'' for both tools.  We propose the Markov Link Method (MLM) as a way to overcome this difficulty.   This method produces consistent estimators that tightly bound the calibration, i.e. the conditional distribution of one tool's measurement given another tool's measurement. It achieves this without any measurement data from both tools applied to the same ``bodies of water.'' Moreover, MLM makes zero assumptions about what calibrations we might expect to see, instead applying a subpopulation-based conditional independence assumption.   We evaluate MLM on a pair of single-cell RNA techniques, obtaining precise calibrations between the tools as well as more accurate models for each tool separately.
\end{abstract}

The modern setting is rife with experimental measurement tools, and it can be very frustrating to understand how the output of these tools relate to one another.  This problem is known as ``calibration'' or ``zeroing'' \cite{bipm2008international}.  A calibration tells us what readings we should expect from one tool, given the reading we obtained from another tool.   Calibration additionally must give uncertainty bounds for how much we can trust those expectations.  Calibration between measurement tools allows us to combine experimental results from different labs and different methodologies into larger scientific theories.

Formally, a calibration is simply a conditional distribution.  We will denote it by $q^*(y|x)$.  As input, this conditional distribution takes the measurement results from one tool on a particular specimen, $x$.  As output, it yields the probability of obtaining result $y$ from a second tool applied to measure the same specimen.  One way to learn the calibration is to measure the same specimens under both tools.  We call this ``joint measurement.''  Unfortunately, calibrations are often required even when joint measurement is unavailable.  For example, if the measurement tool alters the specimen being measured, joint measurement is simply impossible.  In other cases, it may be expensive or impractical.

@@

@@

To make these ideas rigorous, let us develop a little bit of notation.  Let us say we are considering the members of a large population.  For example, each ``specimen`` in the large population might be a human cell or a piece of metal which needs to be tested.  We will assume there are three basic properties of interest for each specimen $i$:

\begin{enumerate}
\item $\ell_i$, the subpopulation or side information.  We assume that the overall population can be split into subpopulations of interest.  For example, we could define subpopulations by looking at cells in different parts of the body, or cells with different sizes.  
\item $X_i$, the result of measuring a specimen with tool I.  For example, perhaps tool I takes a picture of the cell with a destructive electron microscopy method.
\item $Y_i$, the result of measuinrg a specimen with tool II.  For example, perhaps tool II measures the RNA expression of the cell with a destructive sequencing method.
\end{enumerate}

We here consider the case that ``joint measurement'' is impossible or impractical.  In terms of the notation above, that means that for any given specimen we can either observe $\ell_i, X_i$ or $\ell_i,Y_i$.  We can never observe $\ell_i,X_i,Y_i$ for any specimen $i$.  Despite this obstacle, we would like to estimate the conditional distribution $q^*(y|x)\triangleq \mathbb{P}(Y_i=y|X_i=x)$.  Under suitable assumptions, we will see that such estimation is indeed possible using the side information $\ell$.  However, this estimation suffers from so-called \emph{identifiability problems}, which we will describe in detail in the next section.  Fortunately, this identifiability error can be quantified, allowing us to determine exactly what our data tells us about the true calibration.  If the identifiability error is large, the Markov Link method suggests future experiments based on new subpopulations which will hone down on the identifiability problem.

\begin{figure}
\hfill{}\includegraphics[width=.7\textwidth]{../images/bigpicture}\hfill{}
\caption{We consider the case that for each specimen there are two ways we might measure it.  We divide the specimens into two groups.  The specimens on the left are measured with technique I and the specimens on the right are measured with technique II.  In the end, we are left wondering: what would have happened if we had measured the specimens on the left with technique II?  Or what if we had measured the specimens on the right with technique I?  The Markov Link Method gives a way to answer this problem using a subpopultaion analysis.  For each specimen, we take $\ell$ to indicate the subpopulation (indicated here by the color of the neuron, i.e.\ yellow or purple), $X$ to indicate the results we would obtain from that specimen if we measured it with method I, and $Y$ to indicate the results we would obtain from that specimen if we meausured it with method II. \label{fig:bigpicture}}
\end{figure}

The key assumption of the Markov Link Method is a conditional independence assumption: that $\mathbb{P}(Y|X,\ell_i)$ is independent of $\ell$, i.e. $\mathbb{P}(Y=y|X=x,\ell)=q^*(y|x)$ for every value of $\ell$.  Intuitively, this signifies that the manner in which $X$ and $Y$ are related is the same for each subpopulation.  If this assumption is not met, then the method presented here is not applicable.  The validity of these assumptions for a given situation should be closely contemplated.

@@

\begin{itemize}
    \item Quality control for manufacturing.   The surest way to test the reliability of a part is to construct a machine that pushes the part until it breaks.  However, how can we test the reliability of the machine that performs the test?  In each test run there will be some variability induced by the machine itself, which induces a measurement error  In practice, some kind of assumptions about part homogeneity are used to approximate this error (cf.\ \cite{de2005gauge}).  However, if we have two testing machines we can use the MLM to obtain a calibration between the machines, even though we can never test the same part with both machines.  This enables us to bound the overall measurement error.  In this case, $\ell$ might indicate the type of a part being tested, $X$ would indicate the reliability of a part as measured by one machine, and $Y$ would indicate the reliability of a part as measured by another machine.  If the error in machine $Y$ is not correlated to the part type $\ell$, then the MLM assumption certainly holds.  Even if the error is correlated, the MLM assumption may still hold.  For example, imagine that the $Y$ error is correlated with the absolute reliability of the part, this may pose no problem because that reliability is measured by $X$.  As long as the correlations are the result of something which is measured in $X$, the conditional independence assumption will hold.

    \item Combining knowledge across experimental modalities: morphology and transcriptomics.  There are different ways to think about the different types of cells in an organism.  A traditional approach is to classify cells based on what they look like (cf.\ \cite{steinman1973identification,bloomfield1982physiological}).  A more modern approach is to assay the cell's transcriptome (cf.\ \citep{tasic2017shared}).  Unfortunately, modern high-resolution cell photography and single-cell sequencing technologies are both destructive.  As a result, we can't get both kinds of data for the same specimens.  For cells native to regions full of diverse cell-types, finding a correspondence is a real problem.  The result is two completely independent classifications of cells, one for each way of looking at the cell.  MLM allows us to estimate the relationship between those two classification systems, yielding a wholistic understanding of the different types of cells.  In this case, $\ell$ might indicate some side information such as where in the body the cell was found, $X$ would indicate the classification of the cell according to its transcriptomics, and $Y$ would indicate the classification of a cell according to its morphology.  We expect that cell morphology is largely a function of cell transcriptomics.  Thus, as long as the $X$ measurement is sufficiently detailed, we expect that any correlations between $Y$ and $\ell$ would be explained by $X$.  That is, the MLM assumption holds.

    \item Cancer treatment efficacy prediction.  Starting from in-vivo human cancers, many cell-lines have been cultured over the years.  These cell cultures live indefinitely on plates.  Many experiments have been performed to see how these cancer cells respond to treatment.  However, if a treatment work on a particular cultured cell-line, what can we say about which kinds of cancer might respond well to that treatment?  Coarse side-information such as original cancer location is often available for both in-vivo and cultured cells, but this is often a surprisingly weak signal.  Cell transcriptomes provides much more specific information about the cancer, and thus, in theory, what treatments might be appropriate (cf.\ \cite{cieslik2018cancer}).  However, we know that cultured cell-lines look quite different from in-vivo cells (cf. \cite{imamura2015comparison,haibe2013inconsistency}).  Moreover, we have very little joint measurement between the human cells which originally gave rise to a cell culture and the cells that survived to beceome the cell line.  The Markov Link method can leverage the common side-information together with separate transcriptome information to produce a fine-grained correspondence between in-vivo and cultured cells.  This correspondence can used to propose new treatments for cancers.  Here $\ell$ might indicate cancer location, $X$ might indicate transcriptomic expression of cultured cells, and $Y$ might indicate transcriptomic expression of in-vivo cells.  As the transcriptomic expression is much more informative than the cancer location, it is plausible that $X$ might be sufficient to explain any correlations between $\ell$ and $Y$.  Thus the MLM assumption may hold.

    \item Text/image correspondence.  Automatic image captioning is an ongoing effort in machine learning (cf.\ \cite{srivastava2018survey}).  There are three types of data available to help develop such algorithms: text-only data, image-only data, and paired-text-and-image data.  Obviously the last kind is the most useful for automatic image captioning, but there is much less of it.  The Markov Link Method suggests one way to use the more plentiful text-only and image-only data.  We can first apply classic machine learning techniques to get coarse labels for both kinds of data.  Using this side-information to identify subpopulations, the MLM can then deduce a fine-grained correspondence between text and images by combining information from across all the subpopulations.  Here $\ell$ would indicate coarse labels such as ``cat'' or ``street scene.''  These labels could be derived from either images or text and can be trained in a supervised fashion.  Then, $X$ would indicate the image and $Y$ would indicate a caption.  Since the caption should be determined by the picture $X$, the MLM assumption may hold.

@@

The main contributions of this paper are summarized here:

\begin{itemize}
\item Posing of the Markov Link Method assumption as an assumption that may make it possible to perform calibrations without joint measurement.
\item An analysis of some identifiability problems posed by the Markov Link Method assumption.  Although it turns out precise identifiability is often impossible, we formulate an asymptotic theory for bounding the magnitude of the this problem.  In particular, although we often cannot hope to exactly determine the calibration $q^*(y|x)\triangleq\mathbb{P}(Y=y|X=x)$, we can estimate a so-called polytope $\hat \Theta$ which captures our uncertainty about $q^*$ due to identifiability problems.
\item The result of the identifiability analysis is a polytope $\hat \Theta$.  We investigate various ways of measuring the extent of this polytope.  We see how it is possible to understand what this ``extent'' means in terms of the identifiability problem.  When the extent is small, we see that the identifiability problem is actually quite minor.
\item We apply the MLM to two single-cell transcriptomic datasets.  The result uncovers an important and unexpected property of one of the experimental methods.  
\end{itemize}

Our identifiability analysis stands on the shoulders of a long history of relating probabilistic assumptions to probabilistic inequalities on unidentifiable parameters.  Indeed, the core idea of this work is to take an assumption (the Markov Link assumption) and use it to place inequality bounds on an unidentifiable quantity (namely $q^*$).  When inequalities are fairly tight on all sides, we see that much can be learned even when what we want isn't identifiable.  Much of the prior literature in this kind of direction comes from research into causality.  For example, in \cite{bonet2001instrumentality} Bonet uses polytopes not unlike the ones seen here to explore whether a variable can be used as an instrument.  The famous Clauser-Horne-Shimony-Holt inequality was designed to help answer causality questions in quantum physics, but it also sheds light on what distributions are consistent with certain assumptions \cite{clauser1969proposed}.  Indeed the physics literature has contributed many key inequalities (cf. \cite{chaves2014inferring}, \cite{kela2017semidefinite}, and the references therein).  Perhaps the closest work to this one would be \cite{makarov1982estimates}, which used two marginal distributions to get bounds on a property of the joint distribution (namely the distribution of the sum).  We advance this approach to a more general-purpose technique, both by using many subpopulations to refine our estimates and by considering the entire space of possible joint distributions instead of simply a particular aspect of the joint.

The remainder of this work proceeds as follows.  First we give a careful consideration to the Markov Link Method assumption and how it allows us to use side information to estimate the calibration.  We also see when this side information will and will not be helpful.  Next, we outline the Markov Link Method and describe a new asymptotic consistency result concerning the identifiability issues.   Finally, we consider a real dataset.  We perform the method and then attempt to criticize the result three ways: through concerns about not having enough data, through concerns about modeling assumptions, and through concerns about identifiability.  The method largely prevails against these criticisms, and reveals an unexpected feature of the two techniques.

\section{Precise problem formulation and discussion of the identifiability problem}

Let us now be completely rigorous.  We shall make three assumptions:

\begin{enumerate}
\item For each specimen $i$, the distribution of $X_i,Y_i|\ell_i$ may be written 
\[
\mathbb{P}(X_i=x,Y_i=y|\ell_i) = p^*(x|\ell_i)q^*(y|x)
\]
This is the central assumption we have discussed at length above.  For convenience we will also introduce the notation
\[
\mathbb{P}(Y=y|\ell) = h^*(y|\ell)=\sum_x p^*(x|\ell) q^*(y|x)
\]
\item Joint measurement is unavailable.  In particular, we will assume we have $n+m$ individual specimens.  Of these, we have observed $\ell_i,X_i$ for $i\in\{1\cdots n\}$ and $\ell_i,Y_i$ for $i\in n+1\cdots n+m$.  
\item $X$ and $Y$ are discrete random variables with finite support (the general concepts here will apply more generally, but we leave it for future work).  If the data is not discrete, we can always make it so by dividing it into suitable bins.  In this simple case, we may summarize all of this information with two matrices:
\[
N^X_{\ell x} = |\{i\leq n:\ \ell_i=\ell,X_i=x\}|  \qquad N^Y_{\ell y} = |\{i>n:\ \ell_i=\ell,Y_i=x\}|
\]
Thus $N^X,N^Y$ are matrices counting the number of each kind of observation for method I and method II respectively. 
\end{enumerate}

@@

\begin{description}
\item[Not-enough-data problems] We don't have infinite data, so we can't hope to exactly determine $p^*,h^*$.
\item[Identifiability problems] Even if we knew $p^*,h^*$ exactly, it is often impossible to know the value of $q^*$.  We can only ever know that it lies somewhere in $\Theta(p^*,h^*)$.  
\end{description}

Let us now take a moment to understand the set $\Theta$.  This set tells us how the things we can estimate (i.e.\ $p^*,h^*$) inform us about what we want to know (i.e.\ $q^*$).  To understand a bit more concretely how this works, it may be helpful to think of $p^*,h^*,q^*$ as matrices.  From this point of view, one aspect of the definition of $\Theta$ can be written more concisely as a matrix equality constraint on $q^*$: if $q \in \Theta(p^*,h^*)$, then $p^*q=h^*$.  The consequences of this equation depend upon whether $p^*$ has a left-pseudoinverse.
\begin{itemize}
    \item If $p^*$ has a left-pseudoinverse then this equation allows us to uniquely determine what we want ($q^*$) in terms of what we know ($p^*,h^*$).  In general, this will happen when the number of subpopulations outnumbers the number of different states that $X$ can take on.  In this case, the problem should be straightforward to solve.   
    \item If $p^*$ does not have a left-pseudoinverse, then there is an identifiability issue: we can never hope to exactly determine $q^*$.  
\end{itemize}

In this paper we will focus almost entirely on the second case.  In this second case there is a genuine identifiability issue;  it is \emph{impossible} to ever determine the true value of $q^*$, regardless of how much data we have.  It therefore becomes of paramount importance to understand the magnitude of this identifiability problem.  In particular, even though we cannot know $q^*$ exactly, can we place bounds on what $q^*$ could be?

To gain intuition, let us consider a few examples.\vspace{.1in}

\begin{example}
Let $\ell \in \{1,2\}$, $X\in \{1,2,3\}$, $Y\in \{1,2\}$, and 
\[
p^{*}=\left(\begin{array}{ccc}
40\% & 50\% & 10\% \\
10\% & 10\% & 80\%
\end{array}\right)\qquad h^{*}=\left(\begin{array}{cc}
20\% & 80\% \\
40\% & 60\%
\end{array}\right)
\]
That is, for example, $p^*_{1,2}=\mathbb{P}(X=2|\ell=1)=50\%$ and $h^*_{2,1}\mathbb{P}(Y=1|\ell=2)=40\%$.  Let us look at a single equation in the system $p^* q^* = h^*$ entailed by the MLM assumption.  For example, $h^*_{2,1}$, we get
\[
0.4 = h^*_{2,1} = \sum_x p^*_{2,x} q^*_{x,1} = .1 q^*_{1,1} + .1 q^*_{2,1} + .8 q^*_{3,1} 
\]
This immediately tells us something about $q^*$.  For example we now know that $q^*_{1,1} = 10(.4 - .1 q^*_{2,1} - 0.8 q^*_{3,1})$.  But it actually tells us much more than that.  Observe that $0\leq q^*_{xy}\leq 1$ for every $x,y$; in other words, probabilities must be positive by definition.  Thus, in particular,
\[
0.4 = .1 q^*_{1,1} + .1 q^*_{2,1} + .8 q^*_{3,1} \leq .2 + .8 q^*_{3,1}
\]
It follows immediately that $q^*_{3,1}\geq .25$.  Each equation in the MLM system yields insights of this kind; together these insights form the constraints that define $\Theta$.  The more different subpopulations you have, the more equations of this kind you will have.  
\end{example}

@@

Let $N^X,N^Y$ denote matrices describing the original observed data. The purpose of the Markov Link Method is to try to use these matrices to estimate $q^*(y|x)=\mathbb{P}(Y=y|X=x)$, the calibration between $X$ and $Y$.  The Markov Link Method proceeds in two main steps:

>>>>

\newcommand{\figgygr}[2]{\includegraphics[width=#1\textwidth]{#2}}
% \newcommand{\figgygr}[2]{picture here}
\newcommand{\figgy}[1]{\begin{figure}\fbox{\begin{minipage}{\textwidth}#1\end{minipage}}\end{figure}}

@@

\title{Markov Link Method for calibrating without joint measurement, including the case of destructive measurements}
\author{Jackson Loper, Osnat Penn, Trygve Bakken, David Blei, Liam Paninski}

@@

\theoremstyle{definition}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}

@@

\begin{abstract}
A proliferation of new experimental tools has left a serious gap: calibration. Two thermometers can be calibrated against each other by simply measuring the same bodies of water with both thermometers, but the problem is much harder for many modern tools.  One common problem is that we do not have measurements from the same ``body of water'' for both tools.  We propose the Markov Link Method (MLM) as a way to overcome this difficulty.   This method produces consistent estimators that tightly bound the calibration, i.e. the conditional distribution of one tool's measurement given another tool's measurement. It achieves this without any measurement data from both tools applied to the same ``bodies of water.'' Moreover, MLM makes zero assumptions about what calibrations we might expect to see, instead applying a subpopulation-based conditional independence assumption.   We evaluate MLM on a pair of single-cell RNA techniques, obtaining a calibration between the tools.
\end{abstract}

The modern setting is rife with experimental measurement tools, and it can be very frustrating to understand how the output of these tools relate to one another.  This problem is known as ``calibration'' or ``zeroing.''  A calibration tells us what readings we should expect from one tool, given the reading we obtained from another tool.   Calibration additionally must give uncertainty bounds for how much we can trust those expectations  \cite{bipm2008international}.  Calibration between measurement tools allows us to combine experimental results from different labs and different methodologies into larger scientific theories.

Formally, a calibration is simply a conditional distribution.  We will denote it by $q^*(y|x)$.  As input, this conditional distribution takes the measurement result $x$ obtained from one tool on a particular specimen.  As output, it yields the probability of obtaining result $y$ from a second tool applied to measure the same specimen.  One way to learn the calibration is to measure the same specimens under both tools.  We call this ``joint measurement.''  Unfortunately, calibrations are often required even when joint measurement is unavailable.  For example, if the measurement tool significantly alters the specimen being measured, joint measurement is simply impossible.  In other cases, it may be expensive or impractical.

@@

\section{Relation to prior work}

Our main goal is to achieve calibration without joint measurement.  The main obstacle is what is known in the statistics world as an `identifiability problem.'  Due to this problem, we simply \emph{cannot} directly estimate the calibration that we are interested in.  The calibration is simply not ``identifiable.''  Nonetheless, all is not lost.  Our main contribution is to show that one can bound the extent of this identifiability problem.  In short, the fundamental problem cannot generally be vanquished, but we can put it in its place.

Our identifiability analysis stands on the shoulders of a long history of turning probabilistic assumptions into bounds on unidentifiable parameters.  The core idea of the MLM is to take an assumption (the `Markov Link assumption,' which we will define shortly) and use it to place bounds on an unidentifiable quantity (namely the calibration $q^*$).  When these bounds are fairly tight on all sides, we see that much can be learned even though what we want isn't identifiable.  Much of the prior literature in this kind of direction comes from research into causality.  For example, in \cite{bonet2001instrumentality} Bonet uses polytopes not unlike the ones seen here to explore whether a variable can be used as an instrument.  The famous Clauser-Horne-Shimony-Holt inequality was designed to help answer causality questions in quantum physics, but it also sheds light on what distributions are consistent with certain assumptions \cite{clauser1969proposed}.  Indeed the physics literature has contributed many key inequalities (cf. \cite{chaves2014inferring}, \cite{kela2017semidefinite}, and the references therein).  Perhaps the closest work to this one would be \cite{makarov1982estimates}, which used two marginal distributions to get bounds on a property of the joint distribution (namely the distribution of the sum).  We advance this approach to a more general-purpose technique, both by using many subpopulations to refine our estimates and by considering the entire space of possible joint distributions instead of simply a particular aspect of the joint.

It is perhaps worth mentioning that the actual theorem presented here has probably been derived before (although we did not find it in our review of the literature).  Other results in the same flavor as the one presented here might also find a practical use for modern problems.  There is a treasure-trove of ideas in the causality literature; this wealth has not yet been brought fully to bear on the challenging and important problems of calibration for modern experimental modalities.  Our primary intention with this article is to bring attention to the practical utility of these kinds of result.

@@

To make our ideas rigorous, let us develop a little bit of notation.  Let us say we are considering the specimens of a large population.  For example, each specimen in the large population might be a human cell.  Each specimen could also be be a piece of metal which needs to be tested.  We will assume there are three basic properties of interest for each specimen $i$:

\begin{enumerate}
\item $\ell_i$, the subpopulation.  We assume that the overall population can be split into subpopulations of interest.  For example, we could define subpopulations by looking at cells in different parts of the body or cells with different sizes.  
\item $X_i$, the result of measuring a specimen with tool I.  For example, perhaps tool I takes a picture of the cell with a destructive electron microscopy method.
\item $Y_i$, the result of measuinrg a specimen with tool II.  For example, perhaps tool II measures the RNA expression of the cell with a destructive sequencing method.
\end{enumerate}

We here consider the case that ``joint measurement'' is impossible or impractical.  In terms of the notation above, that means that for any given specimen we can either observe $\ell_i, X_i$ or $\ell_i,Y_i$.  We can never observe $\ell_i,X_i,Y_i$ for any specimen $i$.  Despite this obstacle, we would like to estimate the conditional distribution $q^*(y|x)\triangleq \mathbb{P}(Y_i=y|X_i=x)$.  To make this estimation at all possible, we need some kind of assumption.

\figgy{
\hfill{}\figgygr{.7}{../images/bigpicture}\hfill{}
\caption{We consider the case that for each specimen there are two ways we might measure it.  We divide the specimens into two groups.  The specimens on the left are measured with technique I and the specimens on the right are measured with technique II.  In the end, we are left wondering: what would have happened if we had measured the specimens on the left with technique II?  Or what if we had measured the specimens on the right with technique I?  The Markov Link Method gives a way to answer this problem using a subpopulation analysis.   \label{fig:bigpicture}}
}

The key assumption of the Markov Link Method is a conditional independence: that $\mathbb{P}(Y|X,\ell)$ is independent of $\ell$, i.e.

\begin{center}
\fbox{\parbox[c]{4in}{
\vspace{.1in}
\begin{center}
\textbf{The Markov Link Method Assumption}\\$\mathbb{P}(Y=y|X=x,\ell)=q^*(y|x)$ for every value of $\ell$.
\end{center}
\vspace{.1in}
}}
\end{center}

Intuitively, this signifies that the manner in which $X$ predicts $Y$ would be the same for each subpopulation.  If this assumption is not met, then the method presented here is not applicable.  The validity of these assumptions for a given situation should be closely contemplated.

@@

\begin{itemize}
    \item Quality control for manufacturing.   The surest way to test the reliability of a part is to construct a machine that pushes the part until it breaks.  However, how can we test the reliability of the machine that performs the test?  In each test run there will be some variability induced by the machine itself, which induces a measurement error.  In practice, some kind of assumptions about part homogeneity are used to approximate this error (cf.\ \cite{de2005gauge}).  However, if we have two testing machines we can use the MLM to obtain a calibration between the machines, even though we can never test the same part with both machines.  This enables us to bound the overall measurement error.  In this case, $\ell$ might indicate the type of a part being tested, $X$ would indicate the reliability of a part as measured by one machine, and $Y$ would indicate the reliability of a part as measured by another machine.  If the error in machine $Y$ is not correlated to the part type $\ell$, then the MLM assumption certainly holds.  Even if the error is correlated, the MLM assumption may still hold.  For example, imagine that the $Y$ error is correlated with the absolute reliability of the part, this may pose no problem if that reliability is adequately measured by $X$.

    \item Combining knowledge across experimental modalities: morphology and transcriptomics.  There are different ways to think about the different types of cells in an organism.  A traditional approach is to classify cells based on what they look like (cf.\ \cite{steinman1973identification,bloomfield1982physiological}).  A more modern approach is to assay the cell's transcriptome (cf.\ \citep{tasic2017shared}).  Unfortunately, modern high-resolution cell photography and single-cell sequencing technologies are both destructive.  As a result, we can't always get both kinds of data for the same specimens.  For cells native to regions full of diverse cell-types, it is thus quite hard to grasp the correspondence between these different kinds of classification systems.  The result is two completely independent classifications of cells, one for each way of looking at the cell.  MLM allows us to estimate the relationship between those two classification systems, yielding a wholistic understanding of the different types of cells.  In this case, $\ell$ might indicate some side information such as where in the body the cell was found, $X$ would indicate a detailed classification of the cell according to its transcriptomics, and $Y$ would indicate a coarser classification of a cell according to its morphology.  We expect that cell morphology is largely a function of cell transcriptomics.  Thus, as long as the $X$ measurement is sufficiently detailed, we expect that any correlations between $Y$ and $\ell$ would be explained by $X$.  That is, the MLM assumption holds.

    \item Cancer treatment efficacy prediction.  Starting from in-vivo human cancers, many cell-lines have been cultured over the years.  These cell cultures live indefinitely on plates.  Many experiments have been performed to see how these cancer cells respond to treatment.  However, if a treatment works on a particular cultured cell-line, what can we say about whether a treatment will work on an actual in-vivo cancer inside a patient?  Coarse side-information such as original cancer location is often available for both in-vivo and cultured cells, but this is often a surprisingly weak signal.  Cell transcriptomes provides much more specific information about the cancer, and thus, in theory, what treatments might be appropriate (cf.\ \cite{cieslik2018cancer}).  However, we know that cultured cell-lines look quite different from in-vivo cells (cf. \cite{imamura2015comparison,haibe2013inconsistency}).   These cell cultures are subject to quite different pressures, due to the fact that they survive on a plate instead of inside a human being.  The Markov Link method can leverage the common side-information together with separate transcriptome information to understand the correspondence between in-vivo and cultured cells.  If a particular drug is effective on a particular cultured cell-line, we can then look at the corresponding in-vivo transcriptomic profile.  If we find human cancers that match this profile, they might be good candidates for further research using this particular drug.  Here $\ell$ might indicate cancer location, $X$ might indicate transcriptomic expression of cultured cells, and $Y$ might indicate transcriptomic expression of in-vivo cells.  As the transcriptomic expression is much more informative than the cancer location, it is plausible that $X$ might be sufficient to explain any correlations between $\ell$ and $Y$.  Thus the MLM assumption may hold.

    \item Text/image correspondence.  Automatic image captioning is an ongoing effort in machine learning (cf.\ \cite{srivastava2018survey}).  There are three types of data available to help develop such algorithms: text-only data, image-only data, and paired-text-and-image data.  Obviously the last kind is the most useful for automatic image captioning, but there is much less of it.  The Markov Link Method suggests one way to use the more plentiful text-only and image-only data.  We can first apply classic machine learning techniques to get coarse labels for both kinds of data.  Using this side-information to identify subpopulations, the MLM can then deduce a fine-grained correspondence between text and images by combining information from across all the subpopulations.  Here $\ell$ would indicate coarse labels such as ``cat'' or ``street scene.''  These labels could be derived from either images or text and can be trained in a supervised fashion.  $X$ would indicate the image and $Y$ would indicate a caption.  If caption is largely determined by the picture $X$, the MLM assumption may hold.

@@

The rest of the paper proceeds as follows:

\begin{itemize}
\item We describe the identifiability problem and describe how our analysis puts this problem in our place.  We give several examples attempting to give intuition as to why the identifiability problem may not be as severe as it first appears.
\item Applying these ideas, we develop the Markov Link Method for calibration without joint measurements.
\item We apply the MLM to two single-cell transcriptomic methods.  The result uncovers some ways in which the methods appear to be well-calibrated and other areas where the calibration may not be as good. 
\end{itemize}

\section{The identifiability problem}

Let us make three assumptions:

\begin{enumerate}
\item For each specimen $i$, the distribution of $X_i,Y_i|\ell_i$ may be written 
\[
\mathbb{P}(X_i=x,Y_i=y|\ell_i) = p^*(x|\ell_i)q^*(y|x)
\]
This is the central assumption we have discussed at length above.  For convenience we will also introduce the notation
\[
\mathbb{P}(Y=y|\ell) = h^*(y|\ell)=\sum_x p^*(x|\ell) q^*(y|x)
\]
\item Joint measurement is unavailable.  In particular, we will assume we have $n+m$ individual specimens.  Of these, we have observed $\ell_i,X_i$ for $i\in\{1\cdots n\}$ and $\ell_i,Y_i$ for $i\in n+1\cdots n+m$.  
\item $X$ and $Y$ are discrete random variables with finite support (the general concepts here will apply more generally, but we leave it for future work; if the data is not discrete, we can always make it so by dividing it into suitable bins).  In this simple case, we may summarize all of our observed data in two matrices:
\begin{align}
N^X_{\ell x} = |\{i\leq n:\ \ell_i=\ell,X_i=x\}|  \qquad N^Y_{\ell y} = |\{i>n:\ \ell_i=\ell,Y_i=x\}|
\label{eq:twomatrices}
\end{align}
Thus $N^X,N^Y$ are matrices counting the number of each kind of observation for method I and method II respectively. 
\end{enumerate}

@@

\begin{enumerate}
\item Not-enough-data problems. We don't have infinite data, so we can't know the exact values of $p^*,h^*$.
\item Identifiability problems. Even if we knew $p^*,h^*$ exactly, it is often impossible to know the value of $q^*$.  We can only ever know that it lies somewhere in $\Theta(p^*,h^*)$.  
\end{enumerate}

This point of view suggests a practical procedure for estimating $q^*$:

\begin{enumerate}
    \item Use traditional methods to develop estimators $\hat h,\hat p$ so that with more and more data we can guarantee that $\hat h\approx h^*$ and $\hat p\approx p^*$ with high probability.  In particular, we will use a robust pseudocount estimator for $p^*$ and use the MLM assumption to get a high-quality estimator for $h^*$ that leverages information from $N^X$ and $N^Y$.  
    \item Then we may suppose that $q^*$ is probably somewhere \emph{near} to $\Theta(\hat p,\hat h)$.  
\end{enumerate}

It is not completely obvious that this will work.  Certainly $\hat p\approx p^*$ and $\hat h\approx h^*$.  Certainly $q^*\in \Theta(p^*,h^*)$.  But does that guarantee that $q^*$ is nearby to the set $\Theta(\hat p,\hat h)$?  Something could go quite wrong in this line of reasoning.  Fortunately, subject to mild assumptions, one can show that nothing goes wrong.  This is the content of our main technical contribution.  It can be found in Appendix \ref{sec:proof}.

Let us now take a moment to understand the set $\Theta$.  This set encodes exactly how it is that the Markov Link assumption enables us to understand something about the calibration $q^*$.  That is, this set tells us how the things we can estimate (i.e.\ $p^*,h^*$) inform us about what we want to know (i.e.\ $q^*$).  To understand a bit more concretely how this works, it may be helpful to think of $p^*,h^*,q^*$ as matrices.  From this point of view, one aspect of the definition of $\Theta$ can be written more concisely as a matrix equality constraint on $q^*$: if $q \in \Theta(p^*,h^*)$, then $p^*q=h^*$.  This equation is in fact a literal mathematical rendering of the Markov Link assumption of conditional independence.  The consequences of this assumption and its corresponding equation depend upon whether $p^*$ has a left-pseudoinverse.
\begin{itemize}
    \item If $p^*$ has a left-pseudoinverse then this equation allows us to uniquely determine what we want ($q^*$) in terms of what we know ($p^*,h^*$).  In general, this will happen when the number of subpopulations outnumbers the number of different states that $X$ can take on.  In this case, the problem should be straightforward to solve.   
    \item If $p^*$ does not have a left-pseudoinverse, then there is an identifiability issue: we can never hope to exactly determine $q^*$.  
\end{itemize}

In this paper we will focus almost entirely on the second case.  In this second case there is a genuine identifiability issue;  $\Theta(p^*,h^*)$ is nontrivial and so it is \emph{impossible} to ever determine the true value of $q^*$, regardless of how much data we have.  All we have is the equality and inequality bounds that arise from our knowledge that $q^* \in \Theta(p^*,h^*)$.

To gain intuition about the equality and inequality bounds that define the set $\Theta$, let us consider a few examples.\vspace{.1in}

\begin{example}
Let $\ell \in \{1,2\}$, $X\in \{1,2,3\}$, $Y\in \{1,2\}$, and 
\[
p^{*}=\left(\begin{array}{ccc}
40\% & 50\% & 10\% \\
10\% & 10\% & 80\%
\end{array}\right)\qquad h^{*}=\left(\begin{array}{cc}
20\% & 80\% \\
40\% & 60\%
\end{array}\right)
\]
That is, for example, $p^*_{1,2}=\mathbb{P}(X=2|\ell=1)=50\%$ and $h^*_{2,1}\mathbb{P}(Y=1|\ell=2)=40\%$.  Let us look at a single equation in the system $p^* q^* = h^*$ entailed by the MLM assumption.  For example, $h^*_{2,1}$, we get
\[
0.4 = h^*_{2,1} = \sum_x p^*_{2,x} q^*_{x,1} = .1 q^*_{1,1} + .1 q^*_{2,1} + .8 q^*_{3,1} 
\]
We thereby obtain an important constraint on what the calibration $q^*$ might be.

This equation also takes on additional significance in light of the inequality bounds we have: $0\leq q^*_{xy}\leq 1$ for every $x,y$.  In other words, probabilities must be between 0 and 1, by definition.  This may seem trivial, but it has important consequences in light of the constraints imposed by the equation $p^*q=h^*$.  In particular, taking another look at that same equation, we see that
\[
0.4 = .1 q^*_{1,1} + .1 q^*_{2,1} + .8 q^*_{3,1} \leq .2 + .8 q^*_{3,1}
\]
It follows immediately that $q^*_{3,1}\geq .25$.  Each equation in the MLM system yields insights of this kind; together these insights form the constraints that define $\Theta$.  The more different subpopulations you have, the more equations of this kind you will have.  
\end{example}

@@

Let us now exactly describe our approach for calibrating in the absence of joint measurement.  As in Equation (\ref{eq:twomatrices}) above, we assume that $X$ and $Y$ are discrete random variables and $N^X,N^Y$ are the matrices describing the  observed data. The purpose of the Markov Link Method is to try to use these matrices to estimate $q^*(y|x)=\mathbb{P}(Y=y|X=x)$, the calibration between $X$ and $Y$.  The Markov Link Method has three main components:
